probKanren: A Simple Probabilistic extension for microKanren

### 2021-09-15 paper/main.tex
Fix labels
More fixes

%%
%% This command is for the conference information
\conference{PLP '21: The 8th Workshop on Probabilistic Logic Programming,
  September 20--27, 2021, Porto, Portugal}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  Probabilistic programming can be conceptually seen as generalisation
  of logic programming where instead of just returning a set of
  answers to a given query, we also return a probability distribution
  over those answers. But many contemporary probabilistic logic
  programming languages implementations are not simple extensions of
  existing logic programming languages but instead involve their own
  unique implementations. Here we introduce probKanren, a simple
  extension to microKanren that transforms it into a probabilistic
  programming language without needing to make any modifications to
  the underlying logic language's search. We use several illustrative
  examples from the probabilistic programming and program synthesis
  literature to demonstrate the practicality of the approach.

@@

But the conceptual simplicity of this sort of generalisation is not reflected
in the complexity of many existing probabilistic logic programming systems.
They often involve making implementing sophisticated algorithms and the underlying
systems are not just implemented on top of existing logic programming systems.

We believe a conceptually simple extension deserves a conceptually
simple implementation to go along with it. We thus contribute a simple
way to extend microkanren a small logic programming DSL such that it
becomes a probabilistic programming language.

@@

To help explain how to use probKanren we introduce the following example:

@@

\begin{minted}{Scheme}
  (run 1000 (q)
   (conj
    (normal 0 3 q)
    (normal q 2 4))]
\end{minted}

@@

There is a rich history of extending logic programming formalisms to
support probabilistic inference. Early systems like
PRISM\cite{sato1997prism} and ProbLog\cite{de2007problog} allowed
associating discrete distributions with facts. Early versions of
ProbLog were also built on top of Prolog matching one of the goals of
our work. Later work\cite{gutmann2010extending} introduces
distribution clauses so that some continuous distributions. Other work
extended these methods further while focusing on efficient exact
inference algorithms like model weight
integration\cite{islam2012inference, belle2015probabilistic}. Our work
is most similar to \cite{gutmann2011magic} except that while they
combine their forward reasoning with an importance sampler we use a
particle cascade instead which can be more sample efficient.

@@

\textsc{microKanren}\cite{10.1145/2989225.2989230, daniel2018reasoned}
is a pure logic programming language embedded in Scheme. The language
consists of a set of terms, a set of goal primitives, and two run
functions to answer queries in the language. The goal primitives
consist of a \texttt{fresh} form for introducing logic variables, a
unification primitive \texttt{==}, a conjunction combinator
\texttt{conj}, a disjunction combinator \texttt{disj}, and ways to
define and apply relations. Further forms are shown in detail in
\Cref{fig:grammar}.

Goal expressions represent a possibly backtracking search
procedure. These goals all take as input some state and output a
stream of possible output states. We run these goals by starting with
an initial state that holds an empty substitution dictionary, and
passing it into the top-level goal expression which then passes it
recursively down to sub-expressions. Encountering \texttt{fresh} adds
a new logical variable to the substitution; \texttt{conj} will apply
the first goal to the state passed in, and then apply the second goal
to the states associated with each resulting steam finally unioning
the results; \texttt{disj} will apply both goals to the same passed in
state and concatenate the resulting steam; and \texttt{==} unifies its
arguments in the context of the current state discarding any streams
which fail to unify.

To handle goal expressions that might diverge, the streams are
expanded in an interleaving\cite{kiselyov2005backtracking} fashion
giving each branch a chance to produce answers. This interleaving search
is sound and complete\cite{rozplokhas2019certified}.

@@

To create \textsc{probKanren} we extend this grammar with distribution
clauses such as \texttt{normal} and \texttt{bern}. Distribution clauses
take as arguments the parameters of the distribution and a last argument
representing a draw from that distribution. For example, in \texttt{(normal 0 1 x)}
means \texttt{x} represents a draw from $\mathcal{N}(0,1)$ distribution.

These are just another type of goal expression. We do not need to add
a notion of probabilistic variables to the language as they can
treated as logic variables constrained in a particular way. The
semantics of the language though does change from a set of answers to
a query to a probability distribution of answers to a query.

We follow the semantics of \cite{staton2016semantics}, where each
language form has a denotational semantics that associates a
measurable function with each form, and an operational semantics that associates
a sampler with each form.

@@

Probabilistic Programming Languages\cite{wood2014new,
  van2018introduction} are a family of domain specific languages for
posing and efficiently solving probabilistic modelling problems. At
their core, all have a way to \texttt{sample} from a probability
distribution and \texttt{observe} data generated from a probability
distribution.

There are many ways to implement inference algorithms for probabilistic
programming languages but methods based on likelihood-weighting and
sequential monte carlo algorithms are the easiest.

@@

Sequential Monte Carlo\cite{chopin2020introduction}(SMC) is an efficient
online way to sample from probabilistic models especially suited for
state-space domains. If we imagine our probabilistic programs as
straight-line programs with no control-flow we can imagine numbering
every sample function $f_1, f_2, \ldots, f_n$ and every observe
function $g_1, g_2, \ldots, g_n$ then our probability density over our
random variables $x_{0:T}$ and observed data $y_{0:T}$ can be defined as:

@@

The simplest way to sample from our target distribution $p(x_{0:T}
\mid y_{0:T})$ is to samples from a sequence of intermediary
distributions $p(x_{0:t} \mid y_{0:t})$ where $t$ goes from $1$ to
$T$. We do this by sampling a population of $N$ particles
$\{x^{(1)}_t, \ldots, x^{(i)}_t, \ldots x^{(N)}_t\}$ from a proposal
distribution $q(x_t \mid x_{0:t-1}, y_{0:t})$, which when combined
with a set of importance weights $\{w^{(1)}_t, \ldots, w^{(i)}_t,
\ldots w^{(N)}_t\}$ lets us approximate each intermediary
distribution.

Thanks to the recurrence relation

@@

\begin{align*}
x_0^{(i)} &\sim p(x_0) \\
w_0^{(i)} &= 1/N
\end{align*}

@@

\begin{align*}
  x_t^{(i)} &\sim q(x_t \mid x^{(i)}_{0:t-1}) \\
  w_t^{(i)} &\propto w_{t-1}^{(i)} \frac{g(y_t \mid x^{(i)}_t)f(x^{(i)}_t \mid x^{(i)}_{t-1})}{q(x^{(i)}_t \mid x^{(i)}_{0:t-1})}
\end{align*}

@@

Unfortunately, SIS is likely to over time lead to most of the weight
in a small fraction of the particles with the rest of the particles
having negligible weight. This is sometimes called \emph{weight degeneracy}. We
address this by replicating the particles with high weight and
dropping the ones with low weight. This is called a resampling step,
and it happens after each round.

\begin{align*}
  a_t^{(i)} &\sim r(w_{t-1}^{(1:N)}) \\
  x_t^{(i)} &\sim q(x_t \mid x^{(a_t^{(i)})}_{0:t-1}) \\
  w_t^{(i)} &\propto w_{t-1}^{(i)} \frac{g(y_t \mid x^{(i)}_t)f(x^{(i)}_t \mid x^{(i)}_{t-1})}{q(x^{(i)}_t \mid x^{(i)}_{0:t-1})}
\end{align*}

The problem with the above algorithm is over time for many particles
$W_k$ is going to become low and that particle will stop being very
informative of the underlying distribution. To mitigate this issue,
each time we encounter an observation we resample our particles. Resampling
effectively removes particles with low weight and duplicates particles
with higher weight by sampling with replacement our existing particles.

We set $r(w) = \text{Categorical}(w)$, which is called multinomial resampling but there are other methods
as well\cite{douc2005comparison}.

@@

Unfortunately, SMC as defined here requires having access to all the
particles at every step in the process. This conflicts with the
functional nature of our implementation. Instead we make use of
Particle Cascades \cite{PaigeWDT14}, remove this barrier allowing every
particle to be resampled asynchronously with the associated weights
being relative to a global running average. % double-check this

@@

We propose to extend \textsc{microKanren} by turning the search into a
SMC sampler. We accomplish this by augmenting each of the search
streams with a set of particles. These particles represent the
empirical distribution of that stream. Each particle has associated
with it a substitution of all the logic and random variables as well
as a weight that is proportional to the likelihood of the
substitution.

@@

Firstly, the arguments of distribution clauses must be
ground. Secondly, a random variable cannot unify with any arithmetic
expression.

An initial set of particles is created from the probabilistic program
when it is first run. When primitives such as \texttt{normal} we first
check is all the parameter terms are grounded and then if the last
argument is fresh, we sample a value for it for each particle and then
add this sampled value along with the associated logic variable to the
substitution associated with that particle. If all the terms are ground,
we treat the primitive like an observation statement and multiply the
weights of each substitution with the likelihood of the observation.

As disjunctions (\texttt{disj}) are encountered, we split evenly the
number of particles allocated to each stream. Whenever we encounter a
unification primitive, we run a resampling step. This helps to prune
low-weight particles and replicate high weight ones.

As an optimisation we may create more particles during resampling
based on a globally stored a counter of the effective sample size of
all particles across all streams.

This extension does not modify the search and streams are managed
exactly as in microkanren. An additional advantage of this is thanks
to the microkanren search being complete, if we generate enough
particles we are guaranteed to recover the true posterior as all paths
of the search space will eventually be explored.

@@

We validate that probKanren is at least as expressive as other probabilistic
logic programming languages by implementing the Friends who Smoke model.

@@

We made a simple to implement extension to \textsc{microKanren} that
let's us support probabilistic inference on both discrete and
continuous distributions. The approach does not require modifying the
underlying search algorithm or touch any of the backtracking code and
comes with a theoretical guarantee that if the underlying search is
complete then the probabilistic extension will require the true
posterior given enough particles.

>>>>

%%
%% This command is for the conference information
\conference{PLP 2021}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  Probabilistic programming can be conceptually seen as generalisation
  of logic programming where instead of just returning a set of
  answers to a given query, we also return a probability distribution
  over those answers. But many contemporary probabilistic logic
  programming languages implementations are not simple extensions of
  existing logic programming languages but instead involve their own
  unique implementations. Here we introduce \textsc{probKanren}, a simple
  extension to microKanren that transforms it into a probabilistic
  programming language without needing to make any modifications to
  the underlying logic language's search. We use several illustrative
  examples from the probabilistic programming and program synthesis
  literature to demonstrate the practicality of the approach.

@@

But the conceptual simplicity of this sort of generalisation is not
reflected in the complexity of many existing probabilistic logic
programming systems.  They often involve making implementing
sophisticated inference algorithms and the underlying systems are not
just implemented on top of existing logic programming systems.

We believe that a conceptually simple extension deserves a
conceptually simple implementation to go along with it. We thus
contribute a simple way to extend microkanren a small logic
programming domain-specific language such that it becomes a
probabilistic programming language.

@@

To help explain how to use \textsc{probKanren} we introduce the following example:

@@

\begin{minted}{Scheme}
  (run 1000 (q)
   (conj
    (normal 0 3 q)
    (normal q 2 4)))
\end{minted}

@@

There is a rich history of extending logic programming formalisms to
support probabilistic inference. Early systems like
PRISM\cite{sato1997prism} and ProbLog\cite{de2007problog} allowed
associating discrete distributions with facts. Early versions of
ProbLog were also built on top of Prolog, matching one of the goals of
our work. Later work\cite{gutmann2010extending} introduces
distribution clauses so that some continuous distributions like
Gaussians can be represented. Other work extended these methods
further while focusing on efficient exact inference algorithms like
weighted model integration\cite{islam2012inference,
  belle2015probabilistic}. Our work is most similar to
\cite{gutmann2011magic}, except that while they combine their forward
reasoning with an importance sampler we use a particle cascade instead
which can be more sample efficient.

@@

\textsc{microKanren}\cite{10.1145/2989225.2989230, daniel2018reasoned}
is a pure logic programming language embedded in Scheme. The language
consists of a set of terms, a set of goal primitives, and two run
functions that execute queries in the language. The goal primitives
consist of a \texttt{fresh} form for introducing logic variables, a
unification primitive \texttt{==}, a conjunction combinator
\texttt{conj}, a disjunction combinator \texttt{disj}, and ways to
define and apply relations. Further forms are shown in detail in
\Cref{fig:grammar}.

Goal expressions represent a possibly backtracking search
procedure. These goals all take as input some state and output a
stream of possible output states. We run these goals by starting with
an initial state that holds an empty substitution dictionary, and
passing it into the top-level goal expression which then passes it
recursively down to sub-expressions. Encountering \texttt{fresh} adds
a new logical variable to the substitution; \texttt{conj} will apply
the first goal to the state passed in, and then apply the second goal
to the states associated with each resulting stream from the first
goal and finally unify the results; \texttt{disj} will apply both
goals to the same input state and concatenate the resulting streams;
and \texttt{==} unifies its arguments in the context of the current
state, discarding any streams which fail to unify.

To handle goal expressions that might diverge, the streams are
expanded in an interleaving\cite{kiselyov2005backtracking} fashion,
giving each branch a chance to produce answers. This interleaving search
is sound and complete\cite{rozplokhas2019certified}.

@@

To create \textsc{probKanren}, we extend this grammar with distribution
clauses such as \texttt{normal} and \texttt{bern}. Distribution clauses
take as arguments the parameters of the distribution and a last argument
representing a draw from that distribution. For example, in \texttt{(normal 0 1 x)}
means \texttt{x} represents a draw from the $\mathcal{N}(0,1)$ distribution.

These are just another type of goal expression. We do not need to add
a notion of probabilistic variables to the language grammar, as they
can treated as logic variables constrained in a particular way. The
semantics of the language though does change from an answer set to a
probability distribution on that answer set.

We follow the semantics of \cite{staton2016semantics}, where each
language form has the denotational semantics that a
measurable function, and an operational semantics of a sampler.

@@

Probabilistic Programming Languages\cite{wood2014new,
  van2018introduction} are a family of domain-specific languages for
posing and efficiently solving probabilistic modelling problems. At
their core, all have a way to \texttt{sample} and \texttt{observe}
data under a probability distribution or Markov kernel.

There are many ways to inference algorithms for probabilistic
programming languages, but methods based on likelihood-weighting and
Sequential Monte Carlo algorithms are the simplest to implement.

@@

Sequential Monte Carlo\cite{chopin2020introduction}(SMC) is an efficient
online way to sample from probabilistic models, which is especially suited for
state-space domains. If we imagine our probabilistic programs as
straight-line programs with no control-flow we can imagine numbering
every sample function $f_1, f_2, \ldots, f_n$ and every observe
function $g_1, g_2, \ldots, g_n$ then our probability density over our
latent variables $x_{0:T}$ and observed data $y_{0:T}$ can be defined as:

@@

The simplest way to sample from our target distribution $p(x_{0:T}
\mid y_{0:T})$ is to sample from a sequence of intermediary
distributions $p(x_{0:t} \mid y_{0:t})$ where $t$ goes from $1$ to
$T$. We do this by sampling a population of $N$ particles
$\{x^{(1)}_t, \ldots, x^{(i)}_t, \ldots x^{(N)}_t\}$ from a proposal
distribution $q(x_t \mid x_{0:t-1}, y_{0:t})$, which when combined
with a set of importance weights $\{w^{(1)}_t, \ldots, w^{(i)}_t,
\ldots w^{(N)}_t\}$ lets us approximate each intermediary
distribution.

Thanks to the recurrence relation:

@@

\begin{align}
x_0^{(i)} &\sim p(x_0) \\
w_0^{(i)} &= 1/N
\end{align}

@@

\begin{align}
  x_t^{(i)} &\sim q(x_t \mid x^{(i)}_{0:t-1}) \\
  w_t^{(i)} &\propto w_{t-1}^{(i)} \frac{f(x^{(i)}_t \mid x^{(i)}_{t-1})g(y_t \mid x^{(i)}_t)}{q(x^{(i)}_t \mid x^{(i)}_{0:t-1})}
\end{align}

@@

Unfortunately, over time SIS is likely to lead to most of the
importance weight in a small fraction of the particles, with the rest
of the particles having negligible weight. This is usually called
\emph{weight degeneracy}. We address this sample efficiency issue in
the standard way by replicating the particles with high weight and
dropping the ones with low weight. This is called a resampling step,
and it happens after each round. Resampling effectively removes
particles with low weight and duplicates particles with higher weight
by sampling with replacement our existing particles.

\begin{align*}
  a_t^{(i)} &\sim r(w_{t-1}^{(1:N)}) \\
  x_t^{(i)} &\sim q(x_t \mid x^{(a_t^{(i)})}_{0:t-1}) \\
  w_t^{(i)} &\propto w_{t-1}^{(i)} \frac{f(x^{(i)}_t \mid x^{(i)}_{t-1})g(y_t \mid x^{(i)}_t)}{q(x^{(i)}_t \mid x^{(i)}_{0:t-1})}
\end{align*}

For each particle $i$, resampling tells us which of the existing
particles $a_i$ will be its ancestor. We set $r(w) =
\text{Categorical}(w)$, which is called multinomial resampling but
there are other methods as well\cite{douc2005comparison}.

@@

Unfortunately, SMC as defined here requires having access to all the
particles at every step in the process. This conflicts with the
functional nature of our implementation. Instead we make use of
Particle Cascades \cite{PaigeWDT14}, removing this barrier allowing every
particle to be resampled asynchronously with the associated weights
being relative to a global running average. % double-check this

@@

We propose to extend \textsc{microKanren} by turning the search into an
SMC sampler. We accomplish this by augmenting each of the search
streams with a set of particles. These particles represent the
empirical distribution of that stream. Each particle has associated
with it a substitution of all the logic and random variables as well
as a weight that is proportional to the likelihood of the
substitution.

@@

Firstly, the arguments of distribution clauses must be
grounded. Secondly, a random variable cannot unify with any arithmetic
expression.

An initial set of particles is created from the probabilistic program
when it is first run. When encountering primitives such as
\texttt{normal} we first check if all the parameter terms are grounded
and then if the last argument is fresh, we sample a value for it for
each particle and then add this sampled value along with the
associated logic variable to the substitution associated with that
particle. If all the terms are ground, we treat the primitive like an
observation statement and multiply the weights of each substitution
with the likelihood of the observation.

As conjunctions (\texttt{conj}) are encountered, all particles in all
output streams from the first goal become part of the initial states
that are each passed to the second goal.

As disjunctions (\texttt{disj}) are encountered, we split evenly the
number of particles allocated to each stream. Whenever we encounter a
unification primitive, we run a resampling step. This helps to prune
particles with low weight and replicate particles with high weight.

As an optimisation, we may create more particles during resampling
based on a globally stored counter of the effective sample size of
all particles across all streams.

This extension does not modify the search and streams are managed
exactly as in \textsc{microKanren}. An additional advantage, due to
the completeness of the \text{microKanren} search being complete, is
that we are guaranteed to recover the true posterior as all paths of
the search space will eventually be explored given enough particles.

@@

We validate that \textsc{probKanren} is at least as expressive as other probabilistic
logic programming languages by implementing the Friends who Smoke model.

@@

We made a simple to implement extension to \textsc{microKanren} that
lets us support probabilistic inference on both discrete and
continuous distributions. The approach does not require modifying the
underlying search algorithm or touching any of the backtracking code and
comes with a theoretical guarantee that if the underlying search is
complete then the probabilistic extension will recover the true
posterior given enough particles.
